|

### 2024-08-20 reproducibility/manuscript/manuscript.qmd
docs(manuscript): refactor extended data and methods links
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

# Use of `\nameref{sec-methods}` etc. to refer to sections by name.
# is only compatible with LaTeX outputs
number-sections: true

@@

      key: suppfig

@@

      caption-prefix: Supplementary Figure
      reference-prefix: Supplementary Fig.

@@

      latex-list-of-description: Supplementary Figure

@@

# Main text {#sec-main}

@@

Pyro&thinsp;-Velocity recasts the velocity estimation problem into a latent
variable posterior probability inference (@sec-methods). The proposed model is
generative and fully Bayesian, with the different parameters considered as
latent random variables. Each parameter is assigned a prior distribution, and
the model is then conditioned on the raw data (both spliced and unspliced
molecule raw counts) to estimate posterior probability distributions and to
enable uncertainty estimation of velocity. Central to the Pyro&thinsp;-Velocity
model is a *shared latent time* $(t)$: a global latent variable placing each
cell at a particular instant of the modeled dynamical process. From this latent
time, the expectation of spliced and unspliced RNA levels for each gene can be
computed knowing gene-specific kinetics parameters (transcriptional $(\alpha)$,
splicing $(\beta)$, and degradation rates ($\gamma$) and two gene-specific
switching times $(t_{0}^{k})$, and gene-specific transcriptional state $k$) and
integrating the RNA velocity differential equations analogously to what proposed
in the dynamical RNA velocity model by Bergen et al.
[@Bergen2020-pj;@Li2021-qa]. However, we innovate the structure and formulation
of the latent time given that our model is set to jointly fit the entire raw
data matrix, not individual genes. We approach the task of estimating the
posterior distribution by restricting the search within a family of a
variational distribution approximating the true posterior, and taking advantage
of the automatic differentiation variational inference (ADVI) capabilities of
Pyro [@Bingham2018-id;@Kucukelbir2016-bk] ([@fig-model]a, @sec-methods).
Notably, our generative Bayesian model enables to sample from the posterior
distributions of the latent and dependent variables, allowing uncertainty
estimation of phase portraits, high dimensional velocity vector field, and
shared latent time. Around this model, we built several visualizations to assess
uncertainty and functions to recover putative cell fate priming genes and to
investigate how the denoised gene expression recapitulates known cell-type
annotations and their developmental order ([@fig-model]b, @sec-methods).
Although several RNA velocity methods
[@Gayoso2024-fz;@Gorin2022-il;@Gu2022-rh;@Cui2022-em;@Gao2022-jq] have been
recently proposed, and many of them have not been peer-reviewed, here we
showcase the advantages and innovations of Pyro&thinsp;-Velocity by comparing it
with the current state-of-the-art dynamical RNA velocity model proposed by
Bergen et al. [@Bergen2020-pj], and implemented in scVelo.

@@

Pyro&thinsp;-Velocity was also able to define well-known developmental cell
hierarchies in a scRNA-seq dataset of E15.5 mouse pancreas
([@fig-model-results]b-e, $n=3,696$ cells comprising $8$ cell types
([@fig-model-results]b, panel 1)), identifying cell trajectories originating
from ductal progenitor cells and culminated in the production of mature Alpha,
Beta, Delta, and Epsilon cells [@Bastidas-Ponce2019-lf]. Though the scVelo and
Pyro&thinsp;-Velocity models have qualitatively comparable vector fields, the
results generated by scVelo were greatly influenced by gene selection,
neighboring cell number for gene expression smoothing, and cannot provide
confidence assessment of lineage directionality ([@suppfig-model-results]a). In
addition, Pyro&thinsp;-Velocity uncertainty visualizations uncovered that
differentiated cells had a larger vector field uncertainty than stem and
progenitor cells, as would be expected for terminally differentiated cell states
where progression and lineage trajectories would not be anticipated.
Pyro&thinsp;-Velocity also recovered a developmental order in close agreement
with what was estimated by Cytotrace (Spearman's $r=0.95, \; \text{p-value} <
10^{-10}$ between latent time and developmental order)[@Gulati2020-xq] - a
state-of-the-art orthogonal method used to predict cell differentiation based on
the number of expressed genes per cell.

We next used Pyro&thinsp;-Velocity to unbiasedly uncover potential cell fate
determinant markers genes within the E15.5 mouse pancreas. To this end, we first
selected the top 300 genes with the best goodness of fit (negative mean absolute
errors across cells). Next, we ranked them based on the positive (or negative)
correlation between inferred shared time and corresponding denoised spliced
expression per cell ([@fig-model-results]d, Methods). This procedure
successfully recovered key developmental genes that are activated (or repressed)
during endocrinogenesis ([@fig-model-results]d-e, @suppfig-model-results). Among
the top positively correlated genes identified by Pyro&thinsp;-Velocity,
endocrine Beta cells expressed insulin 1 and 2 (Ins1, Ins2) and Nnat, which is a
key regulator Beta cell insulin generation [@Millership2018-vd] and Ppp1r1a
which is a marker of Beta cells [@Jiang2013-pt]. The ranking procedure proposed
by scVelo failed to identify Ins1, Ins2, Nnat, and Ppp1r1a as lineage-restricted
genes in their top 50 ranked gene list. The inability of scVelo to identify
these key developmental genes could be likely explained by the kNN smoothing of
read count and filtering steps (Supplementary Table 1, 2). Visual inspection of
these four top genes by Pyro&thinsp;-Velocity revealed a single activation phase
during cell type transitions based on phase portrait curves. We also propose
their visualization through ``rainbow plots'' to illustrate denoised single gene
spliced expression across shared time and to investigate the inferred cell order
based on known annotations. Further, UMAP embedding of cells colored based on
denoised gene expression confirmed the cell fate markers' smooth and progressive
activation (or repression) towards endocrine cells ([@fig-model-results]e,
[@suppfig-model-results]).

Ground truth velocity is not a quantity we can easily assess. Therefore, no
robust and quantitative benchmarking of RNA velocity methods has been proposed.
However, recent single-cell assays allow recording lineage relationships and can
be used to assess the ability of different methods in recapitulating cell fate
progression. Therefore, we next systematically compared scVelo and
Pyro&thinsp;-Velocity ability to recover single cell fates in Lineage And RNA
RecoverY (LARRY) barcoded hematopoietic cells. LARRY leverages unique lentiviral
barcodes inserted into 3' untranslated region (UTR) of GFP reporter that is
integrated as a single copy into parental cells. Then progeny can be clonally
traced to follow the descendent cell fates over time. LARRY has been recently
applied to in vitro differentiation of human blood and accurately recovered
expected cell fate and lineage trajectories [@Weinreb2020-mv]. This dataset
sampled differentiation over the course of 2, 4, and 6 days and contained
$49,302$ cells that each could be traced with at least one barcode. Because
LARRY couples single-cell transcriptomics and direct cell lineage tracing, this
data set provides a unique opportunity to benchmark RNA Velocity quantitatively
and to provide insights into the correct interpretation of recovered velocity
vector fields and latent cell times. To quantitatively assess the agreement
between observed and predicted fates, we considered two metrics: 1. average
cosine similarity (cs) between vectors from the RNA Velocity vector field and
the clonal progression vector field, i.e., a vector field derived directly from
the lineage data ([@sec-methods]), 2. Spearman's correlation $(\rho)$ between
cell latent (shared) time and fate potency scores derived from Cytotrace and
Cospar. Cospar, our gold standard, is a state-of-the-art method designed
explicitly for predicting fate potency based on LARRY data, while Cytotrace
leverages only transcriptomic data. We selected cell state changes associated
with monocyte and neutrophil lineages because both lineages have the largest
numbers of cells in the scRNA sequencing data set and well-known cell fate
trajectories [@Petukhov2018-nr]. As expected, the Pyro&thinsp;-Velocity vector
field accurately predicted cell state trajectories from hematopoietic progenitor
cells into mature monocytes or neutrophils and had high cosine similarity
between the inferred vector field and clonal progression vector field assigned
by LARRY (monocytes $cs=0.49$, neutrophils $cs=0.38$).  High Spearman's
correlation was also seen between shared latent time and Cospar fate potency
(monocytes $\rho=0.60$, neutrophils $\rho=0.74$) or shared latent time with
Cytotrace fate potency (monocytes $\rho=0.925$, neutrophils $\rho=0.925$). By
contrast, scVelo generated vector fields with reversed order or low correlation
(cosine similarity: $-0.38$, monocytes, cosine similarity: $0.32$, neutrophils).
Moreover, Cospar fate potency scores were negative ($\rho=-0.59$, monocytes) or
non-significantly correlated ($\rho=-0.07$, neutrophils).  A similar low
correlation was observed for scVelo and Cytotrace fate potency scores (monocytes
$\rho=0.309$, neutrophils $\rho=0.751$) ([@fig-lineage-tracing]a-b,
[@suppfig-lineage-tracing]c).

Next, we evaluated the ability of the Pyro&thinsp;-Velocity models to recover
bi-fate or multi-fate trajectories by considering clones belonging to monocyte
and neutrophil lineages or using the entire dataset. Not surprisingly, both
methods showed their limitations, given that they are not designed to model and
capture multiple fate trajectories ([@suppfig-lineage-tracing-2]c-d). The
recovered vector fields and cell latent (shared) time show inconsistent
directions for both methods. However, Pyro&thinsp;-Velocity identified higher
uncertainty for some of the problematic regions. Furthermore, we discovered that
vector field and shared time estimation could be significantly improved by
extending the original kinetic model by allowing each gene a time lag for the
transcriptional activation and a baseline expression level potentially different
than 0 (Pyro&thinsp;-Velocity Model 2, [@sec-methods])
([@fig-lineage-tracing]c,d, [@suppfig-graphical-models]b,c). This extended model
recapitulated a vector field consistent with the clonal progression vector field
for most branches with some spurious predictions within the intermediate cell
state between stem cells and Megakaryocytes and most terminal Neutrophils
(bi-fate cosine similarity: $0.62$, multi-fate cosine similarity: $0.63$).
However, the proposed uncertainty estimation accurately flagged these
subpopulations to warn users of potential false discoveries. Notably, this
extended model significantly improved the correlation of cell shared time with
the clonal fate potency (bi-fate $\rho=0.65$, multi-fate $\rho=0.74$) and also
improved the correlation with Cytotrace fate potency (bi-fate $\rho=0.931$,
multi-fate $\rho=0.614$). Remarkably, the shared time per cell from
Pyro&thinsp;-Velocity also outperformed Cytotrace for developmental order
prediction ([@suppfig-lineage-tracing]a,c). Importantly, this extended model
retained its ability to accurately identify cell lineage trajectories and shared
time for all the other datasets analyzed above ([@suppfig-model-2-result]).
Taken together, our approach using LARRY barcoded datasets established the first
quantitative benchmark of RNA velocity-based cell fate trajectory inference and
provides an exciting new framework for continued optimization of multi-fate
predictions in the future.

In conclusion, Pyro-Velocity provides the first probabilistic, interpretable,
end-to-end Bayesian inference framework for RNA Velocity that learns cell
dynamics and recovers cell fate from multimodal raw read counts.  This approach
provides uncertainty estimations of cell fate trajectories and novel
visualization approaches to mitigate false discoveries. Pyro-Velocity recovered
shared time and vector fields that are recapitulated in true developmental order
and outperformed state-of-the-art methods like Cytotrace and scVelo
([@suppfig-lineage-tracing]a).

# Software availability

We developed our Pyro-Velocity models with Pyro (version 1.6.0) probabilistic
programming language and PyTorch (version 1.8.1). Pyro-Velocity source code is
available at <https://github.com/pinellolab/pyrovelocity>. The version of
Pyro-Velocity (v0.1.0) used to generate the results presented in this manuscript
has been deposited on Zenodo: <https://zenodo.org/record/7072876#.Yx-2e-zMLxg>.

# Acknowledgments

# Author contributions

# Competing financial interests statement

We assume the dynamical gene expression is determined by the RNA splicing
process, and infer the unspliced and spliced gene expression level from the
differential equations proposed in velocyto [@La_Manno2018-lj] and scVelo
[@Bergen2020-pj]

\begin{align}
\frac{d u\left(\tau^{\left(k_{cg}\right)}\right)}{d \tau^{\left(k_{cg}\right)}}
  &= \alpha^{\left(k_{cg}\right)}-\beta_g u\left(\tau^{\left(k_{cg}\right)}\right), \label{eq-dudt}\\
\frac{d s\left(\tau^{\left(k_{cg}\right)}\right)}{d \tau^{\left(k_{cg}\right)}}
  &= \beta_g u\left(\tau^{\left(k_{cg}\right)}\right)-\gamma_g s\left(\tau^{\left(k_{cg}\right)}\right). \label{eq-dsdt}
\end{align}

# References

@@

# Supplementary Figures

::: {#suppfig-graphical-models}
![](figures/fig.pdf)

@@

::: {#suppfig-model-results}
![](figures/fig.pdf)

@@

::: {#suppfig-model-2-result}
![](figures/fig.pdf)

Extension of @fig-model-results, Pyro&thinsp;-Velocity (Model 2) outperforms
scVelo on PBMC and pancreas datasets. Same format as in @fig-model-results and
@suppfig-model-results.
:::

::: {#suppfig-lineage-tracing}
![](figures/fig.pdf)

@@

::: {#suppfig-lineage-tracing-2}
![](figures/fig.pdf)

@@

![An example figure (an empty plot)](figures/fig.pdf){#suppfig-plot}

@@

# Supporting Information {.appendix}

>>>>

# Use of `\nameref{sec-methods}` versus [Methods](#sec-methods) or [@sec-methods]
# to refer to sections by name is only compatible with LaTeX outputs
number-sections: false

@@

      key: extfig

@@

      caption-prefix: Extended Data Fig.
      reference-prefix: Extended Data Fig.

@@

      latex-list-of-description: Extended Data Figure

@@

# Main text {#sec-main .unnumbered}

@@

Pyro&thinsp;-Velocity recasts the velocity estimation problem into a latent
variable posterior probability inference ([Methods](#sec-methods)). The proposed
model is generative and fully Bayesian, with the different parameters considered
as latent random variables. Each parameter is assigned a prior distribution, and
the model is then conditioned on the raw data (both spliced and unspliced
molecule raw counts) to estimate posterior probability distributions and to
enable uncertainty estimation of velocity. Central to the Pyro&thinsp;-Velocity
model is a *shared latent time* $(t)$: a global latent variable placing each
cell at a particular instant of the modeled dynamical process. From this latent
time, the expectation of spliced and unspliced RNA levels for each gene can be
computed knowing gene-specific kinetics parameters (transcriptional $(\alpha)$,
splicing $(\beta)$, and degradation rates ($\gamma$) and two gene-specific
switching times $(t_{0}^{k})$, and gene-specific transcriptional state $k$) and
integrating the RNA velocity differential equations analogously to what proposed
in the dynamical RNA velocity model by Bergen et al.
[@Bergen2020-pj;@Li2021-qa]. However, we innovate the structure and formulation
of the latent time given that our model is set to jointly fit the entire raw
data matrix, not individual genes. We approach the task of estimating the
posterior distribution by restricting the search within a family of a
variational distribution approximating the true posterior, and taking advantage
of the automatic differentiation variational inference (ADVI) capabilities of
Pyro [@Bingham2018-id;@Kucukelbir2016-bk] ([@fig-model]a,
[Methods](#sec-methods)). Notably, our generative Bayesian model enables to
sample from the posterior distributions of the latent and dependent variables,
allowing uncertainty estimation of phase portraits, high dimensional velocity
vector field, and shared latent time. Around this model, we built several
visualizations to assess uncertainty and functions to recover putative cell fate
priming genes and to investigate how the denoised gene expression recapitulates
known cell-type annotations and their developmental order ([@fig-model]b,
[Methods](#sec-methods)). Although several RNA velocity methods
[@Gayoso2024-fz;@Gorin2022-il;@Gu2022-rh;@Cui2022-em;@Gao2022-jq] have been
recently proposed, here we showcase the advantages and innovations of
Pyro&thinsp;-Velocity by comparing it with the current state-of-the-art
dynamical RNA velocity model proposed by Bergen et al. [@Bergen2020-pj], and
implemented in scVelo.

@@

Pyro&thinsp;-Velocity was also able to define well-known developmental cell
hierarchies in a scRNA-seq dataset of E15.5 mouse pancreas
([@fig-model-results]b-e, $n=3,696$ cells comprising $8$ cell types
([@fig-model-results]b, panel 1)), identifying cell trajectories originating
from ductal progenitor cells and culminated in the production of mature Alpha,
Beta, Delta, and Epsilon cells [@Bastidas-Ponce2019-lf]. Though the scVelo and
Pyro&thinsp;-Velocity models have qualitatively comparable vector fields, the
results generated by scVelo were greatly influenced by gene selection,
neighboring cell number for gene expression smoothing, and cannot provide
confidence assessment of lineage directionality ([@extfig-model-results]a). In
addition, Pyro&thinsp;-Velocity uncertainty visualizations uncovered that
differentiated cells had a larger vector field uncertainty than stem and
progenitor cells, as would be expected for terminally differentiated cell states
where progression and lineage trajectories would not be anticipated.
Pyro&thinsp;-Velocity also recovered a developmental order in close agreement
with what was estimated by Cytotrace (Spearman's $r=0.95, \; \text{p-value} <
10^{-10}$ between latent time and developmental order)[@Gulati2020-xq] - a
state-of-the-art orthogonal method used to predict cell differentiation based on
the number of expressed genes per cell.

We next used Pyro&thinsp;-Velocity to unbiasedly uncover potential cell fate
determinant markers genes within the E15.5 mouse pancreas. To this end, we first
selected the top 300 genes with the best goodness of fit (negative mean absolute
errors across cells). Next, we ranked them based on the positive (or negative)
correlation between inferred shared time and corresponding denoised spliced
expression per cell ([@fig-model-results]d, Methods). This procedure
successfully recovered key developmental genes that are activated (or repressed)
during endocrinogenesis ([@fig-model-results]d-e, @extfig-model-results). Among
the top positively correlated genes identified by Pyro&thinsp;-Velocity,
endocrine Beta cells expressed insulin 1 and 2 (Ins1, Ins2) and Nnat, which is a
key regulator Beta cell insulin generation [@Millership2018-vd] and Ppp1r1a
which is a marker of Beta cells [@Jiang2013-pt]. The ranking procedure proposed
by scVelo failed to identify Ins1, Ins2, Nnat, and Ppp1r1a as lineage-restricted
genes in their top 50 ranked gene list. The inability of scVelo to identify
these key developmental genes could be likely explained by the kNN smoothing of
read count and filtering steps (Supplementary Table 1, 2). Visual inspection of
these four top genes by Pyro&thinsp;-Velocity revealed a single activation phase
during cell type transitions based on phase portrait curves. We also propose
their visualization through ``rainbow plots'' to illustrate denoised single gene
spliced expression across shared time and to investigate the inferred cell order
based on known annotations. Further, UMAP embedding of cells colored based on
denoised gene expression confirmed the cell fate markers' smooth and progressive
activation (or repression) towards endocrine cells ([@fig-model-results]e,
[@extfig-model-results]).

Ground truth velocity is not a quantity we can easily assess. Therefore, no
robust and quantitative benchmarking of RNA velocity methods has been proposed.
However, recent single-cell assays allow recording lineage relationships and can
be used to assess the ability of different methods in recapitulating cell fate
progression. Therefore, we next systematically compared scVelo and
Pyro&thinsp;-Velocity ability to recover single cell fates in Lineage And RNA
RecoverY (LARRY) barcoded hematopoietic cells. LARRY leverages unique lentiviral
barcodes inserted into 3' untranslated region (UTR) of GFP reporter that is
integrated as a single copy into parental cells. Then progeny can be clonally
traced to follow the descendent cell fates over time. LARRY has been recently
applied to in vitro differentiation of human blood and accurately recovered
expected cell fate and lineage trajectories [@Weinreb2020-mv]. This dataset
sampled differentiation over the course of 2, 4, and 6 days and contained
$49,302$ cells that each could be traced with at least one barcode. Because
LARRY couples single-cell transcriptomics and direct cell lineage tracing, this
data set provides a unique opportunity to benchmark RNA Velocity quantitatively
and to provide insights into the correct interpretation of recovered velocity
vector fields and latent cell times. To quantitatively assess the agreement
between observed and predicted fates, we considered two metrics: 1. average
cosine similarity (cs) between vectors from the RNA Velocity vector field and
the clonal progression vector field, i.e., a vector field derived directly from
the lineage data ([Methods](#sec-methods)), 2. Spearman's correlation $(\rho)$
between cell latent (shared) time and fate potency scores derived from Cytotrace
and Cospar. Cospar, our gold standard, is a state-of-the-art method designed
explicitly for predicting fate potency based on LARRY data, while Cytotrace
leverages only transcriptomic data. We selected cell state changes associated
with monocyte and neutrophil lineages because both lineages have the largest
numbers of cells in the scRNA sequencing data set and well-known cell fate
trajectories [@Petukhov2018-nr]. As expected, the Pyro&thinsp;-Velocity vector
field accurately predicted cell state trajectories from hematopoietic progenitor
cells into mature monocytes or neutrophils and had high cosine similarity
between the inferred vector field and clonal progression vector field assigned
by LARRY (monocytes $cs=0.49$, neutrophils $cs=0.38$).  High Spearman's
correlation was also seen between shared latent time and Cospar fate potency
(monocytes $\rho=0.60$, neutrophils $\rho=0.74$) or shared latent time with
Cytotrace fate potency (monocytes $\rho=0.925$, neutrophils $\rho=0.925$). By
contrast, scVelo generated vector fields with reversed order or low correlation
(cosine similarity: $-0.38$, monocytes, cosine similarity: $0.32$, neutrophils).
Moreover, Cospar fate potency scores were negative ($\rho=-0.59$, monocytes) or
non-significantly correlated ($\rho=-0.07$, neutrophils).  A similar low
correlation was observed for scVelo and Cytotrace fate potency scores (monocytes
$\rho=0.309$, neutrophils $\rho=0.751$) ([@fig-lineage-tracing]a-b,
[@extfig-lineage-tracing]c).

Next, we evaluated the ability of the Pyro&thinsp;-Velocity models to recover
bi-fate or multi-fate trajectories by considering clones belonging to monocyte
and neutrophil lineages or using the entire dataset. Not surprisingly, both
methods showed their limitations, given that they are not designed to model and
capture multiple fate trajectories ([@extfig-lineage-tracing-2]c-d). The
recovered vector fields and cell latent (shared) time show inconsistent
directions for both methods. However, Pyro&thinsp;-Velocity identified higher
uncertainty for some of the problematic regions. Furthermore, we discovered that
vector field and shared time estimation could be significantly improved by
extending the original kinetic model by allowing each gene a time lag for the
transcriptional activation and a baseline expression level potentially different
than 0 (Pyro&thinsp;-Velocity Model 2, [Methods](#sec-methods))
([@fig-lineage-tracing]c,d, [@extfig-graphical-models]b,c). This extended model
recapitulated a vector field consistent with the clonal progression vector field
for most branches with some spurious predictions within the intermediate cell
state between stem cells and Megakaryocytes and most terminal Neutrophils
(bi-fate cosine similarity: $0.62$, multi-fate cosine similarity: $0.63$).
However, the proposed uncertainty estimation accurately flagged these
subpopulations to warn users of potential false discoveries. Notably, this
extended model significantly improved the correlation of cell shared time with
the clonal fate potency (bi-fate $\rho=0.65$, multi-fate $\rho=0.74$) and also
improved the correlation with Cytotrace fate potency (bi-fate $\rho=0.931$,
multi-fate $\rho=0.614$). Remarkably, the shared time per cell from
Pyro&thinsp;-Velocity also outperformed Cytotrace for developmental order
prediction ([@extfig-lineage-tracing]a,c). Importantly, this extended model
retained its ability to accurately identify cell lineage trajectories and shared
time for all the other datasets analyzed above ([@extfig-model-2-result]). Taken
together, our approach using LARRY barcoded datasets established the first
quantitative benchmark of RNA velocity-based cell fate trajectory inference and
provides an exciting new framework for continued optimization of multi-fate
predictions in the future.

In conclusion, Pyro&thinsp;-Velocity provides the first probabilistic,
interpretable, end-to-end Bayesian inference framework for RNA Velocity that
learns cell dynamics and recovers cell fate from multimodal raw read counts.
This approach provides uncertainty estimations of cell fate trajectories and
novel visualization approaches to mitigate false discoveries.
Pyro&thinsp;-Velocity recovered shared time and vector fields that are
recapitulated in true developmental order and outperformed state-of-the-art
methods like Cytotrace and scVelo ([@extfig-lineage-tracing]a).

# References {.unnumbered}

@@

\twocolumn

We assume the dynamical gene expression is determined by the RNA splicing
process, and infer the unspliced and spliced gene expression level from the
differential equations proposed in velocyto [@La_Manno2018-lj] and scVelo
[@Bergen2020-pj]
\begin{align}
\frac{d u\left(\tau^{\left(k_{cg}\right)}\right)}{d \tau^{\left(k_{cg}\right)}}
  &= \alpha^{\left(k_{cg}\right)}-\beta_g u\left(\tau^{\left(k_{cg}\right)}\right), \label{eq-dudt}\\
\frac{d s\left(\tau^{\left(k_{cg}\right)}\right)}{d \tau^{\left(k_{cg}\right)}}
  &= \beta_g u\left(\tau^{\left(k_{cg}\right)}\right)-\gamma_g s\left(\tau^{\left(k_{cg}\right)}\right). \label{eq-dsdt}
\end{align}

## Model evaluation and selection {#sec-model-selection}

Pyro&thinsp;-Velocity provides two model formulations as described above.
Users can compare their predictive performance to decide which of the two models
to use for a given dataset. Namely, the mean absolute errors of posterior
samples prediction of spliced and unspliced raw read counts $x_{icg}$ from its
observation $\tilde{X}_{cg}$ per gene and cell combinations
\begin{align*}
\frac{1}{N_i N_c N_g} \sum_{i=1}^{N_i} \sum_{c=1}^{N_c} \sum_{g=1}^{N_g} |x_{icg} - \tilde{X}_{cg}|,
\end{align*}
where $i$ is the posterior sample index, $N_i$ is the posterior sample number,
$N_c$ is the cell number, $N_g$ is the gene number. Then it is possible to
evaluate the cell fate inference performance of the models based on two metrics:

1. the consistency between the velocity vector field and the *clonal progression
vector field* using grid-based cosine similarity; 2. Spearman’s correlation of
velocity-based shared (latent) time with Cospar fate potency and the silver
standard Cytotrace score. Both metrics are implemented in scipy.

# Code availability

We developed our Pyro&thinsp;-Velocity models with Pyro (version 1.6.0)
probabilistic programming language and PyTorch (version 1.8.1).
Pyro&thinsp;-Velocity source code is available at
<https://github.com/pinellolab/pyrovelocity>. The version of
Pyro&thinsp;-Velocity (v0.1.0) used to generate the results presented in this
manuscript has been deposited on Zenodo:
<https://zenodo.org/record/7072876#.Yx-2e-zMLxg>.

<!-- # End notes {.unnumbered} -->

# Acknowledgments {.unnumbered}

# Author contributions {.unnumbered}

# Competing financial interests statement {.unnumbered}

# Extended Data Figures

::: {#extfig-graphical-models}
![](figures/fig.pdf)

@@

::: {#extfig-model-results}
![](figures/fig.pdf)

@@

::: {#extfig-model-2-result}
![](figures/fig.pdf)

Extension of @fig-model-results, Pyro&thinsp;-Velocity (Model 2) outperforms
scVelo on PBMC and pancreas datasets. Same format as in @fig-model-results and
@extfig-model-results.
:::

::: {#extfig-lineage-tracing}
![](figures/fig.pdf)

@@

::: {#extfig-lineage-tracing-2}
![](figures/fig.pdf)

@@

<!-- ![An example figure (an empty plot)](figures/fig.pdf){#extfig-plot} -->

@@

<!-- # Supporting Information {.appendix} -->
